---
title: "Neotoma-charcoal workshop"
author: "Petr Kuneš & Thomas Giesecke"
date: "2023-02-15"
output:
  html_document:
    toc: yes
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We will introduce you to the [Neotoma Paleoecology Database](www.neotomadb.org) environment, including explorer, [Tilia](https://www.neotomadb.org/apps/tilia) and the `neotoma2` R package. The database uses an application programming interface (API), which serves the explorer as well as Tilia but also lets you query the database in R. We will show you which information you can most easily extract in either of these choices and provide a hands-on R exercise on working with charcoal and pollen data.

[The `neotoma2`package](https://github.com/NeotomaDB/neotoma2) is available from GitHub and can be installed in R using the `devtools` package. We also load other packages.

```{r install packages, echo=TRUE, message=FALSE, warning=FALSE}
devtools::install_github('NeotomaDB/neotoma2')
library(neotoma2)
library(riojaPlot)
library(rioja)
library(vegan)
```

## Using single site

### Downloading charcoal and pollen data

As an example, we will use the site 'Prášilské jezero', a mountain lake located in Czechia (Šumava Mts.). This site has been analyzed for several proxies, including pollen and macrocharcoal.

First, we obtain the `siteid` and other metadata including its datasets for the chosen site and plot its position:

```{r datasets code}
PRA_ds <- get_datasets(get_sites(sitename = "Prášilské jezero"), all_data = TRUE)
plotLeaflet(PRA_ds)
```

Filter out the datasets 'pollen' and 'charcoal'

```{r filter datasets}
PRA_pol_ch <- PRA_ds %>% neotoma2::filter(datasettype %in% c("pollen", "charcoal"))
```

And pull in and extract the sample data

```{r pull samples, results='hide'}
PRA_rec <- PRA_pol_ch %>% get_downloads()
PRA_samp <- PRA_rec %>% samples()
```

### Filter samples

Now we will `filter` only `charcoal` samples with values as `area` (function `filter` of `dplyr` package). We also select only relevant columns, and merge `variablename` and `elementtype`. 

```{r filter charcoal}
PRA_charcoal <- samples(PRA_rec) %>% dplyr::filter(ecologicalgroup == "CHAR" & elementtype == "area") %>%
  dplyr::select(depth, age, variablename, elementtype, value) %>%
  dplyr::mutate(variable = paste(variablename, elementtype, sep = " "), .keep = "unused") %>%
  dplyr::group_by(depth, age) %>%
  dplyr::arrange(depth)
```

In the next step, we pull in pollen samples that are measured using `NISP` (number of identified specimen), only pollen element (to avoid e.g., stomata). Finally, we `filter` taxa to calculate pollen percentages for ecological groups of trees and shrubs (`TRSH`) and upland herbs (`UPHE`).

```{r pull pollen samples}
PRA_samp_short <- samples(PRA_rec) %>% 
  dplyr::filter(ecologicalgroup %in% c("TRSH","UPHE")) %>%
  dplyr::filter(elementtype == "pollen") %>%
  dplyr::filter(units == "NISP")
```

### Data transformation

Finally, we need to transform the values to proportions using all TRSH and HERB taxa.

```{r proportions}
PRA_pollen_perc <- PRA_samp_short %>%
  dplyr::group_by(depth, age) %>%
  dplyr::mutate(pollencount = sum(value, na.rm = TRUE)) %>%
  dplyr::group_by(variablename) %>% 
  dplyr::mutate(prop = value / pollencount) %>% 
  dplyr::arrange(desc(age))
```

For further use in other packages, it is useful to pivot the data to a "wide" table, with `variablenames` as column headings. We do this for both pollen and charcoal datasets using `tidyr`.

```{r pivot}
PRA_pollen_perc <- PRA_pollen_perc %>%
  dplyr::select(depth, age, variablename, prop) %>% 
  dplyr::mutate(prop = as.numeric(prop))

PRA_pollen_wide <- tidyr::pivot_wider(PRA_pollen_perc,
                             id_cols = c(depth, age),
                             names_from = variablename,
                             values_from = prop,
                             values_fill = 0)

PRA_charcoal_wide <- tidyr::pivot_wider(PRA_charcoal,
                                        id_cols = c(depth, age),
                                        names_from = variable,
                                        values_from = value,
                                        values_fill = 0)
```

### Stratigraphic plotting

To plot a stratigraphic diagram, we can use `riojaPlot` and combine pollen and charcoal proxies. For pollen, we select 15 most frequent TRSH `taxa` to be plotted.

```{r}
PRA_plot_taxa <- taxa(PRA_rec) %>% dplyr::filter(ecologicalgroup %in% c("TRSH")) %>%
  dplyr::filter(elementtype == "pollen") %>%
  dplyr::arrange(desc(samples)) %>% 
  head(n = 15) 
```

#### Zonation

Now, we calculate a constrained hierarchical cluster analysis using `CONISS` (constrained incremental sum-of-squares cluster analysis). We further determine the significant zones (clusters) by the broken stick model (`bstick` - 7 splits / 8 zones).

```{r cluster}
PRA_clust <- chclust(vegdist(PRA_pollen_wide[,-1:-2], "euclidean"), method = "coniss")
plot(PRA_clust, hang = -1)
bstick(PRA_clust)
PRA_zones<-cutree(PRA_clust,7)
```

We plot the stratigraphic diagram with riojaPlot, and add determined zones.

```{r}
PRA_plot1 <- riojaPlot(PRA_pollen_wide[,-1:-2]*100, PRA_pollen_wide[,1:2],
          selVars = PRA_plot_taxa$variablename,
          scale.percent = TRUE,
          sec.yvar.name="age",
          plot.sec.axis = TRUE,
          srt.xlabel = 60,
          xRight = 0.8)
PRA_plot2 <- addRPClust(PRA_plot1, PRA_clust, xLeft=0.8, xRight = 0.9)
addRPClustZone(PRA_plot2, PRA_clust, nZone=7, xRight = 0.9, col = "red")
PRA_plot3 <- riojaPlot(PRA_charcoal_wide[,-1:-2], PRA_charcoal_wide[,1:2], 
          minmax = data.frame(0,6),
          plot.poly = FALSE,
          plot.bar = TRUE,
          plot.line = FALSE,
          lwd.bar = 2,
          col.bar = "black",
          xLeft = 0.9,
          riojaPlot = PRA_plot2)
```

To further adapt your stratigraphic plot, please refer to [riojaPlot gallery](To%20further%20adapt%20your%20stratigraphic%20plot,%20please%20refer%20to%20riojaPlot%20gallery%20https://nsj3.github.io/riojaPlot_gallery/riojaPlot_Gallery.html#1).

## Comparing the trend in charcoal to variability in pollen composition

### Show the trend in charcoal deposition

First, we plot log-transformed charcoal area data on a time axis. Then we add different trends: `lowess` (locally-weighted polynomial regression); `loess` (fits a polynomial surface determined by one or more numerical predictors, using local fitting); `smooth.spline` (fits a cubic smoothing spline).

```{r trends}
plot(PRA_charcoal_wide$age, log(PRA_charcoal_wide$`Charcoal area`+1), ylim=c(0, 6), xlab = "Age cal BP", ylab = "CHAR")
lines(PRA_charcoal_wide$age, lowess(PRA_charcoal_wide$`Charcoal area`, f=0.1)$y, col="red")
lines(PRA_charcoal_wide$age, loess(PRA_charcoal_wide$`Charcoal area`~PRA_charcoal_wide$age, span = 0.2)$fitted, col="blue")
lines(PRA_charcoal_wide$age, smooth.spline(PRA_charcoal_wide$`Charcoal area`, df = 25)$y, col="green")
legend(1,6, legend = c("LOWESS", "LOESS", "smooth spline"), col = c("red", "blue", "green"), lty=1, cex=0.8)
```

To estimate the values for given depths/ages in the model, we need to apply `predict()`

```{r predict}
PRA_char_loess <- loess(PRA_charcoal_wide$`Charcoal area`~PRA_charcoal_wide$age, span = 0.20)
PRA_char_pred <- stats::predict(PRA_char_loess, PRA_pollen_wide$age)
```

### Multivariate analysis (ordination)

#### Principle component analysis

We apply the Principle component analysis (PCA), here `rda()`, a linear unconstrained ordination method. As the method uses Euclidean distance to calculate dissimilarity between samples, we need to transform the data first to prevent the double-zero problem by applying e.g., the Hellinger transformation (square-root) using `decostand`. Next, we plot the ordination using the `biplot` function, adding a fitted environmental variable (charcoal) with `envfit()`.

```{r pca}
PRA_pca <- rda(decostand(PRA_pollen_wide[,-1:-2], method = "hellinger"))
PRA_PCA_plot <- biplot(PRA_pca)
plot(envfit(PRA_pca, data.frame(CHAR = PRA_char_pred)))
orditorp(PRA_PCA_plot, display = "species")
```

Now we fit a smooth surface of charcoal variable to the ordination plot:

```{r surface}
PRA_ordisurf <- ordisurf(PRA_pca, PRA_char_pred, knots = 2)
```

The ordination plots are not ideal, especially with many samples and variables. Therefore, we may play around with different settings to produce a nicer ordination plot. First, we color samples according to their zones, then select a few taxa to display the main gradients. Finally, we plot over the smooth surface plot with charcoal.

```{r nice plot}
tax.sc<-scores(PRA_pca, display="species", choice=1:2)
tax<-c("Pinus", "Betula", "Picea", "Fagus", "Poaceae", "Corylus")

ordiplot(PRA_pca, type = "n")
points(PRA_pca, col = PRA_zones + 6, pch = PRA_zones, lwd = 2)
for (i in tax) {arrows(0,0,tax.sc[i,1], tax.sc[i,2], length = 0.1, angle = 20, code = 2, lwd = 2, col = "red")
  text(tax.sc[i,1]+0.05, tax.sc[i,2]+0.05, i , cex = 1.2, col = "red")
  }
plot(PRA_ordisurf, add=TRUE)
legend(-1, 0, legend=c("zone 1", "zone 2", "zone 3", "zone 4", "zone 5", "zone 6", "zone 7"),
       pch = c(1,2,3,4,5,6,7), pt.cex = 1.2, pt.lwd=2, bty = "n", col=c(7,8,9,10,11,12,13))
```

#### Detrended correspondence analysis

Here we perform the Detrended correspondence analysis (DCA) with `decorana`. Correspondence analysis is a weighted averaging method. DCA detrends the first ordination axis by segments (hammer effect) often used to determine the extent of compositional turnover (4 SD units = complete turnover) and decide whether the species response are unimodal or linear. DCA also attempts to remove the arch effect from ordination (compare the PCA and DCA ordination plots).

Check the gradient length - look at the axis length of DCA1

```{r dca}
PRA_dca <- decorana(decostand(PRA_pollen_wide[,-1:-2], method = "hellinger"))
PRA_dca
plot(PRA_dca, display = "sites")
```

### Check correlation between charcoal and pollen taxa

We will use two additional packages, allowing faster calculation and better results plotting.

```{r install correlation, echo=TRUE, message=FALSE, warning=FALSE}
library(corrplot)
library(Hmisc)
```

First, we merge the data and log transform charcoal. Then we calculate the correlation matrix using rcorr. Then we plot the significant results (p \< 0.01).\

```{r correlation}
PRA_cor_data <- merge(PRA_pollen_wide, PRA_charcoal_wide)
PRA_cor_data$`Charcoal area` <- log(PRA_cor_data$`Charcoal area` + 1)
PRA_cor <- rcorr(as.matrix(PRA_cor_data[,-1:-2]), type = "pearson")

cor.data <- as.matrix(PRA_cor$r[89,which(PRA_cor$P[89,-89]<0.05)])
colnames(cor.data) <- "Charcoal"

corrplot(t(cor.data))
```
